Preregistration for modelling in ecolgy and conservation

### 2021-05-17 EcoConsPreReg-supervisor_draft-Sep-2020.Rmd
updates to ms and add supervisor draft from sep 2020

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elliot Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
output: 
  distill::distill_article:
    self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

**Abstract:** *Ecology is at risk of a 'reproducibility crisis', in particular, there are serious transparency problems in the reporting of ecological models, which is concerning because a lack of complete and transparent reporting practices provide the conditions in which questionable research practices are likely to arise, leading to irreproducible findings. Irreproducibility of model-based research threatens the progress of both scientific discovery in 'pure' research, and risks wrong or sub-optimal conservation and management decisions being implemented when models are used in decision-support. Preregistration is an open-science practice adopted in scientific disciplines that have confronted the reproducibility crisis, which aims to improve research transparency and mitigate questionable research practices. There has been little uptake of preregistration in ecology and conservation, and within disciplines wherein preregistration has been adopted and become common practice, many modellers and researchers engaged in non-hypothesis testing research have vehemently eschewed preregistration on the grounds that existing preregistration templates are irrelevant to their research because they are focussed on null-hypothesis significance testing. We took a user-centred design approach to the task of translating preregistration into model-based research contexts for ecology and conservation. Structured discussion and collaborative activities underpinned the development of a preregistration template that captures critical analysis decisions in the process of the model development cycle. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the linear procedure of preregistration in which a researcher shifts from ideation to analysis planning, to analysis and then reporting. We propose a methodology for implementing 'adaptive preregistration' leveraging git and GitHub, which accommodates common practices in modelling where some analysis decisions are difficult or cannot be made in advance, where downstream decisions might be dependent on the outcome of earlier decisions, and where preliminary analyses might need to be conducted before proceeding with further modelling. We tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia. This research provides an extensible template and methodology for implementing adaptive preregistration of ecological models, to be used in either pure or applied settings. We hope that this research will facilitate wider uptake of preregistration in ecology and that our proposed methodology for implementing adaptive preregistration might be applied in model-based research within science more broadly. Modellers need no longer cry "but, I can't preregister my research."*

**Keywords** *(no more than 8 for MEE).***:** *preregistration, transparency, reproducibility, modelling, ecology, conservation, open-science, research software engineering.*

@@

We should be concerned about research transparency because, a lack of complete and transparent reporting practices provide the conditions in which QRPs are likely to arise [@Fidler:2017he]. Moreover, as Morrison [-@Morrison:2016cd] demonstrated, when models are not reported transparently with sufficient detail, they cannot be reproduced. This is particularly problematic because in ecology, models are typically used in decision-support, often directly underpinning policy and management decisions. Incomplete and non-transparent reporting of the model and the modelling process risk the model being used in inappropriate applications or decision contexts [@Grimm:2014es] because the underlying modelling choices and assumptions are not communicated [@deVos:2011]. Alternatively, a lack of transparency may mask poor model design or serious flaws in the model, with the risk that they be used in decision support resulting in wrong, or irreversible decisions [@Grimm:2014es;@Schmolke:2010fd]. A lack of transparency in reporting also risks undermining the credibility of the model, such that the decision-maker rejects the use of the model and instead turns to unstructured and subjective judgments [@Grimm:2014es;@Addison:2013wa], thus contributing further to "decision complacency", or the "knowing-doing" gap [@Gardner:2018dm].

@@

Ecology has seen recent improvements in journal policy concerning reporting guidelines and data/code archiving, but there has been slow uptake of other important tools, like preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines to think that preregistration is "not for them", this was a particularly salient sentiment for researchers engaged in non-experimental studies. We have experienced this sentiment ourselves when looking at preregistration templates to use for model-based research --- nearly all preregistration templates are geared towards experimental hypothesis-testing research. Such templates contain items that are simply irrelevant to model-based and/or applied ecological research. More importantly, they do not reflect the iterative nature of model development [@MacEachern2019] and fail to capture critical analytic decisions in the modelling process. Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however it is steeped heavily in an an academic psychology context, is focussed on the use of models for hypothesis-testing, neglects the entire model development process, and approaches modelling from a very narrow epistemic function, in a way that is not commonly applied in ecology.

Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology. Different research methodologies, and their specific and nuanced application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology.

@@

@@

1\. Define the goals of the resource under development and identify the audience,

2\. Characterise users and their needs,

3\. Design mock-ups of your product,

4\. Build prototypes of selected mock-ups and test with users,

5\. Implement the product, write technical implementations and build the product.

@@

We define the primary purpose of the template as "to delimit researcher degrees of freedom" when conducting model-based research in applied ecology and conservation. Although others have defined the purpose of preregistration to include 'delineating between *exploratory* from *confirmatory* research' \<refs\>, this is not the intended purpose of the template here. Beyond hypothesis-testing contexts, this is an idealised and simplifying dichotomy that, in practice, is more complicated and nuanced. For instance researchers might move back and forth between different reasoning-modes throughout the modelling cycle. Thinking through how modes of inference might hold within applied ecological modelling contexts requires further, more careful thought, and consideration of perspectives from philosophy of science. Consequently, we wish to deemphasise the purpose of distinguishing between exploratory and confirmatory research, and focus the design of the templates in being able to delimit researcher degrees of freedom, so as to reduce the potential for cognitive biases, such as confirmation bias, and other questionable research practices, like cherry-picking, HARKing etc. \<insert ref to my QRP manuscript\>.

@@

**1\. Identify Scientific Workflows for ecological modelling, both real/in-practice and idealised.**

@@

**2\. Translating Workflows into Preregistration Templates**

@@

1\. Decision Phase - each has a "title" and "description" or "definition", steps are grouped under phases. \<Note, that these are being scrapped from the template and are turned into tool-tips\>

2\. Decision Step - each has a "title" and "description" or "definition" choices are grouped into 'steps', which corresponds to an activity or set of related activities in a scientific workflow. \<again, these are being scrapped and turned into tool-tips.\>

3\. Choices ---\> each 'choice' corresponds to a uniquely numbered item on the PRT requiring a response from user. The style should be in the form of a directive or a question. E.g. "explain how you will do xyzzy". Or "What performance measure will you use for assessing 'goodness of fit'?"

**3\. Designing a new methodology for implementing preregistration in model-based research**

@@

Modellers \<here, here, and here\> have expressed concern about preregistration making modelling a more difficult task than it already is, and hampering the ability to meet project deadlines, especially in decision-problems. Many modellers already struggle with sufficiently documenting models during publication and reporting \<ref, model reporting and transparency review\>, and modellers working in government or other agencies have expressed concern that preregistration will simply add to the list of  'red-tape' expected of them. Given these concerns, and considering the concerns of many with regard to the iterative nature of model development precluding making preregistration impossible or inappropriate, we figured it was of utmost importance that the templates be also evaluated on their ease-of-use and feasibility.

@@

@@

1\. Opinionated Design - norms around modelling practice.  'Good Modelling Practice' Literature. This also has the benefit of aiding new modellers.

2\. A related point: linguistic ambiguity and other problems with modelling concepts and terms. Contested terms in the ecological modelling literature about what each different step is, their activities, etc.

3\. *research transparency*. We felt that the decision-points captured in the template are critical to increasing reporting transparency sufficiently for others to properly evaluate the modelling and analysis.

@@

@@

------ *We did this by:*

*1\. Identifying norms and practices to develop 'idealised' / 'generalised' research workflows that describe sets of commonly undertaken tasks (and some sort of order to reflect the workflow), within an ecological modelling research analysis (within applied settings??). (idealised + practical).*

*2\. We then transformed these workflows into a 'preregistration template', by distilling a set of decision-items that, we think, should a) adequately constrain researcher degrees of freedom to mitigate the risk of Questionable Research Practices in an ecological modelling exercise and b) adequately facilitates* research/analytic transparency. We transformed the decision-steps (the highest resolution) into template items posed as directives or questions for preregistration users to complete, as appropriate, for their study's preregistration.

@@

3\. Our UX approach revealed critical features of the model development process, namely the inherent iterative nature of model development, and requirements for model checking that forced us to reconsider the procedure for completing a preregistration suitable to ecological modelling. Consequently, we took an 'expanded view of preregistration', "Adaptive Preregistration", as proposed and first described by Srivastava. Adaptive preregistration

@@

4\.

@@

5\. Furthermore, we created an interface for completing and amending interim preregistrations using shinyApps to facilitate uptake of preregistration in both pure and applied ecological modelling research.

*consisting of a series of*

@@

We encourage modellers to preregister their research using our template and methodology here, and to provide feedback and suggestions for improvement. Our template + shiny app is hosted on \<insert GitHub link\>, suggested changes may be submitted in the form of a pull-request to the repository \<link to how to make a pull request\>, or by posting a new issue \<link to creating a new issue on the repo\>. The repository and issues are publicly available, so any discussion will be transparent and open to the ecological modelling community.  
--------------------------
# References

>>>>

---
title: "Preregistration for modelling in ecolgy and conservation"
description: |
  Working Draft.
author:
  - name: Elise Gould 
    affiliation: School of BioSciences, University of Melbourne
  - name: Chris Jones
    affiliation: Department of Environment Land and Water
  - name: Libby Rumpff
    affiliation: School of BioSciences, University of Melbourne
  - name: Hannah Fraser
    affiliation: School of BioSciences, University of Melbourne
  - name: Fiona Fidler
    affiliation: School of BioSciences, University of Melbourne
date: "`r Sys.Date()`"
bibliography: ./data/references.bib
output: word_document
  #distill::distill_article:
    #self_contained: false
---
**Intended journal:** Methods in Ecology and Evolution.

@@

**Abstract:**

**1.** *Preregistration is an open-science practice adopted in scientific disciplines that have begun to confront the 'reproducibility crisis', which aims to improve research transparency and mitigate questionable research practices within a study. There has been little uptake of preregistration in ecology and conservation, and many modellers and researchers engaged in non-hypothesis testing research both within and outside of ecology have eschewed preregistration on the grounds that existing templates are irrelevant because they are focussed on null-hypothesis significance testing.*

*2. We advocate that preregistration can and should be used in model-based research in ecology and conservation but that preregistration should reflect the norms and practice of the research context in which it is applied in order to adequately restrict researcher degrees of freedom that are unique to a domain-specific methodology. We took a user-centred design approach to the task of translating preregistration into model-based research contexts for ecology and conservation.*

*3. We identified a generalised 'modelling workflow' that captures critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we propose a methodology for implementing 'adaptive preregistration' leveraging git and GitHub. We tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia.*

*4. This research provides a template and methodology for implementing adaptive preregistration of ecological models in either pure or applied settings, and will facilitate the wider uptake of preregistration in ecology. Although we focus on ecology and conservation in this paper, the template is extensible to alternative fields and methods, and the proposed methodology for implementing adaptive preregistration can be applied in model-based research within science more broadly. Modellers need no longer cry "but, I can't preregister my research."*

**Keywords:** *preregistration, transparency, reproducibility, modelling, ecology, conservation, open-science, research software engineering.*

@@

A lack of complete and transparent reporting practices is concerning because it  provides the conditions in which QRPs are likely to arise [@Fidler:2017he]. Moreover, as Morrison [-@Morrison:2016cd] demonstrated, when models are not reported transparently with sufficient detail, they cannot be reproduced. This is particularly problematic because in ecology, models are typically used in decision-support, often directly underpinning policy and management decisions. Incomplete and non-transparent reporting of the model and the modelling process risk the model being used in inappropriate applications or decision contexts [@Grimm:2014es] because the underlying modelling choices and assumptions are not communicated [@deVos:2011]. Alternatively, a lack of transparency may mask poor model design or serious flaws in the model, with the risk that they be used in decision support resulting in wrong, or irreversible decisions [@Grimm:2014es;@Schmolke:2010fd]. A lack of transparency in reporting also risks undermining the credibility of the model, such that the decision-maker rejects the use of the model and instead turns to unstructured and subjective judgments [@Grimm:2014es;@Addison:2013wa], thus contributing further to "decision complacency", or the "knowing-doing" gap [@Gardner:2018dm].

@@

Ecology has seen recent improvements in journal policy concerning reporting guidelines and data/code archiving, but there has been slow uptake of other important tools, like preregistration and registered reports (Parker, Fraser, and Nakagawa 2019). Pu et al. 2019 found that conceptual barriers may prevent users from different disciplines to think that preregistration is "not for them", this was a particularly salient sentiment for researchers engaged in non-experimental studies. We have experienced this sentiment ourselves when looking at preregistration templates to use for model-based research --- nearly all preregistration templates are geared towards experimental hypothesis-testing research. Such templates contain items that are simply irrelevant to model-based and/or applied ecological research. More importantly, they do not reflect the iterative nature of model development [@MacEachern2019] and fail to capture critical analytic decisions in the modelling process. Presently, we are aware of only a single preregistration template for modelling [@Cruwell2019b], however it is steeped heavily in an an academic psychology context, is focussed on the use of models for hypothesis-testing, neglects the8 model development process, and approaches modelling from a very narrow epistemic function, in a way that is not commonly applied in ecology.

Preregistration cannot simply be ported over to other disciplines and adopted as-is. If we wish to adopt preregistration in ecology, specifically within model-based research, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology. Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology.

@@

@TODO: add argument from Berna's formal theory paper about the need for testing and evaluating new reforms. While only a qualitative assessment... still useful information, such as XYZ.

Mention testing / eval here? Yes need to mention a little bit, BUT perhaps could save Berna's quote for discussion.

@@

1. Define the goals of the resource under development and identify the audience,

2. Characterise users and their needs,

3. Design mock-ups of your product,

4. Build prototypes of selected mock-ups and test with users,

5. Implement the product, write technical implementations and build the product.

@@

We define the primary purpose of the template as "to delimit researcher degrees of freedom" when conducting model-based research in applied ecology and conservation.  Although the purpose of preregistration is predominantly conceptualised in terms of distinguishing between exploratory from confirmatory research [ e.g. @Nosek:2018eg], we wish to emphasise that this is not the intended purpose of this template.  We adopt a more general stance towards the function of preregistration, i.e. that preregistration should mitigate against questionable research practices by reducing the potential for cognitive biases to cause opportunistic misuse of researcher degrees of freedom, such as cherry-picking and HARKing.

Explanation of why we adopt this stance (problem with exploratory vs. confirmatory distinction):

*Beyond hypothesis-testing contexts, this is an idealised and simplifying dichotomy that, in practice, is more complicated and nuanced. For instance researchers might move back and forth between different reasoning-modes throughout the modelling cycle. Thinking through how modes of inference might hold within applied ecological modelling contexts requires further, more careful thought, and consideration of perspectives from philosophy of science.*

@@

**1. Identify Scientific Workflows for ecological modelling, both real/in-practice and idealised.**

@@

**2. Translating Workflows into Preregistration Templates**

@@

1. Decision Phase - each has a "title" and "description" or "definition", steps are grouped under phases. \<Note, that these are being scrapped from the template and are turned into tool-tips\>

2. Decision Step - each has a "title" and "description" or "definition" choices are grouped into 'steps', which corresponds to an activity or set of related activities in a scientific workflow. \<again, these are being scrapped and turned into tool-tips.\>

3. Choices ---\> each 'choice' corresponds to a uniquely numbered item on the PRT requiring a response from user. The style should be in the form of a directive or a question. E.g. "explain how you will do xyzzy". Or "What performance measure will you use for assessing 'goodness of fit'?"

**3. Designing a new methodology for implementing preregistration in model-based research**

@@

Modellers [@MacEachern2019] \<here, here, and here\> have expressed concern about preregistration making modelling a more difficult task than it already is, and hampering the ability to meet project deadlines, especially in decision-problems. Many modellers already struggle with sufficiently documenting models during publication and reporting \<ref, model reporting and transparency review\>, and modellers working in government or other agencies have expressed concern that preregistration will simply add to the list of  'red-tape' expected of them. Given these concerns, and considering the concerns of many with regard to the iterative nature of model development precluding making preregistration impossible or inappropriate, we figured it was of utmost importance that the templates be also evaluated on their ease-of-use and feasibility.

@@

Berna - should evaluate open science reforms.

@@

1. Opinionated Design - norms around modelling practice.  'Good Modelling Practice' Literature. This also has the benefit of aiding new modellers.

2. A related point: linguistic ambiguity and other problems with modelling concepts and terms. Contested terms in the ecological modelling literature about what each different step is, their activities, etc.

3. *research transparency*. We felt that the decision-points captured in the template are critical to increasing reporting transparency sufficiently for others to properly evaluate the modelling and analysis.

@@

## Case Study ##

Should speak to the following, which could be included in the supporting materials.

\- Completed preregistration

\- Interim Preregistration Documents

\- link to repository

\- screenshots illustrating  how the GitHub process is used.

@@

*------ We did this by (overview of what we did):*

*1. Identifying norms and practices to develop 'idealised' / 'generalised' research workflows that describe sets of commonly undertaken tasks (and some sort of order to reflect the workflow), within an ecological modelling research analysis (within applied settings??). (idealised + practical).*

*2. We then transformed these workflows into a 'preregistration template', by distilling a set of decision-items that, we think, should a) adequately constrain researcher degrees of freedom to mitigate the risk of Questionable Research Practices in an ecological modelling exercise and b) adequately facilitates* research/analytic transparency. We transformed the decision-steps (the highest resolution) into template items posed as directives or questions for preregistration users to complete, as appropriate, for their study's preregistration.

@@

3. Our UX approach revealed critical features of the model development process, namely the inherent iterative nature of model development, and requirements for model checking that forced us to reconsider the procedure for completing a preregistration suitable to ecological modelling. Consequently, we took an 'expanded view of preregistration', "Adaptive Preregistration", as proposed and first described by Srivastava. Adaptive preregistration

@@

4.

@@

5. Furthermore, we created an interface for completing and amending interim preregistrations using shinyApps to facilitate uptake of preregistration in both pure and applied ecological modelling research.

***Evaluation and testing (I think this provides a nice lead on to the guts of the discussion, so it should be at the end of our summary of what we did).***

Berna Devezer et al. argue that most open science reforms propose them and are never formally tested. In this research, we provide a qualitative assessment of our new method only.

Thus preregistration is based on the working hypothesis that it is able to mitigate against QRPs.

We need empirical testing of this hypothesis, also so we can empirically evaluate whether the new method meets its intended purpose (link back to intended purpose of the preregistration specified in the methods section).

Then, following that, we need metaresearch, so empirical research at the level of the body of literature, perhaps examining transparency is the easiest starting point.. but how would you empirically evaluate a series of models??)

Either way alternative benefits of preregistration (pu et al. talks about this.. what about others).

@@

We encourage modellers to preregister their research using our template and methodology here, and to provide feedback and suggestions for improvement. Our template + shiny app is hosted on \<insert GitHub link\>, suggested changes may be submitted in the form of a pull-request to the repository \<link to how to make a pull request\>, or by posting a new issue \<link to creating a new issue on the repo\>. The repository and issues are publicly available, so any discussion will be transparent and open to the ecological modelling community.

**Putting a positive spin on the time-costs of recording all decisions. The result of completing an adaptive preregistration using our methodology is that you get a 'Lab Notebook' at the end of it.**

People have described modelling as being a digital or computational computer lab. I think the term '*in silico* experiments' has been used to describe some computational modelling, especially for simulation modelling. I'm pretty sure that some authors have made arguments that we would expect rigorous note-taking and recording in the lab environment, so why should we expect any less in a computational modelling setting?

**Preregistration will NOT lead to punch card style analysis (ala MacEachern).**

In our preregistration testing with the case study, we found the opposite, that it prompted critical thinking about analytical choices during modelling.

Moreover, our design template is *opinionated*. (See Elith's latest SDM reporting guideline) -\> it provides a *guide*. For those whom are at risk of 'punch card style analysis', ala MacEachern.  
--------------------------
# References
